---
title: "Practical Machine Learning Course Project"
author: "Sue Chew"
date: "October 24, 2015"
output: html_document
---

# Overview

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. We are given data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants, from [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har). They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The goal is to predict how well they did the exercise.

## Data preparation

Download and load data.
```{r eval=FALSE}
library(caret)

if (!file.exists('data')) { dir.create('data') }
if (!file.exists('data/pml-training.csv') || !file.exists('data/pml-testing.csv')) {
    download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv', destfile='data/pml-training.csv')
    download.file('https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv', destfile='data/pml-testing.csv')
}

training = read.table('data/pml-training.csv', sep=',', header=TRUE, na.strings=c('', 'NA', '#DIV/0!'))
# Validation test set will be used to predict the out of sample error.
validation = read.table('data/pml-testing.csv', sep=',', header=TRUE, na.strings=c('', 'NA', '#DIV/0!'))
```

The following data cleaning were performed:

* Removed covariates with NA's
* Removed covariates with near zero values
* Removed columns that are not accelerometer measurements, such as `user_name`.

```{r eval=FALSE}
# Drop columns with NA's
training <- training[ , sapply(training, Negate(anyNA)), drop=FALSE]

# Obtain covariates with near zero values.
nsz <- nearZeroVar(training[ , -which(names(training) == "classe")], saveMetrics=TRUE)
dim(subset(nsz, nzv==TRUE))
# Remove them from the training set.
training <- training[, -which(names(training) %in% row.names(subset(nsz, nzv==TRUE)))]

# Remove columns not related to measurements
training <- training[, -which(names(training) %in% c('X', 'user_name', 'raw_timestamp_part_1', 'raw_timestamp_part_2', 'cvtd_timestamp', 'num_window'))]
```


## Model building

The training data was partitioned to `training.train` to use for training the mode, and `training.test` to test out of sample accuracy.

```{r eval=FALSE}
set.seed(3456)
inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
training.train <- training[inTrain,]
training.test <- training[-inTrain,]
```

Random forest with 5-fold cross validation was used to fit a model.

```{r eval=FALSE}
# Specify 5-fold cross validation
modFit <- train(classe ~ ., data=training.train, method="rf",
                trControl=trainControl( method = "repeatedcv", number = 5))
# save the model
saveRDS(modFit, "rfModel.rds")
# Show a summary of the model.
summary(modFit)
```

The model was used to predict on `training.test` in order to determine out of sample error. A confusion matrix is shown comparing predicted values of `classe` to the actual values.

```{r eval=FALSE}
testPredictions <- predict(modFit, newdata=training.test)
confusionMatrix(testPredictions, training.test$classe)
```
The error rates in the confusion matrix are estimates of the out of sample errors, that is, an accuracy of 0.9974 or Kappa of 0.9933.

# Run on validation set and write answers
```{r eval=FALSE}
validationPredictions <- predict(modFit, newdata=validation)

if (!file.exists('answers')) { dir.create('answers') }
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("answers/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(validationPredictions)

```

